State of the Practice for Lattice Boltzmann Method Software

### 2022-02-25 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Minor edits throughout, sections 1 to 8 are mostly done

\begin{keyword}
	Lattice Boltzmann Method (LBM), scientific computing, software engineering,
	software quality, Analytic Hierarchy Process
\end{keyword}

@@

A small sample of important applications of LBM include the following: designing
fuel cells \citep{ZhangEtAl2018}, modelling groundwater flow
\citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
system \citep{SadeghiEtAl2020}.  As these examples illustrate, LBM can be used
for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety.  Given the important applications of LBM,
users of the libraries will be concerned with software qualities like
reliability, robustness, reproducibility, performance, correctness and
verifiability.  Since their time is valuable, developers that create or modify
LBM libraries will have additional quality concerns, including maintainability,
reusability, and understandability.  With the quality considerations in mind,
and the potentially overwhelming number of choices for existing LBM libraries,
the time is right to assess the state of the practice.  The goal of this report
is to analyze the current state of LBM software development and compare it to
the development practices employed for research software in general.  We want to
highlight success stories that LBM developers can share amongst their
community, and with the broader research software community, while at the same time watching for areas for potential future improvement.

@@

Only so much information can be gleaned by looking at software repositories.  To
gain additional insight, we need to interview developers.  We need to learn
their concerns, how they deal with these concerns; we need to learn what pain
points exist for them.  We wish to know what practices are used by the top LBM
projects, so that others can potentially emulate these practices. We also wish
to identify new practices by borrowing successful ideas from other domains. For
LBM developers interested in future cutting edge practices, ideas that are
proposed, but not yet in practice, should also be identified.  The above points
are covered by the questions outlined below:

@@

We developed a methodology for evaluating the state of the practice of
scientific software \citep{SmithEtAl2021}. This methodology can be instantiated
for a specific domain of scientific software, which in the current case is LBM
software.  Since we group best practices around conventional software qualities,
the first section below provides definitions for the qualities of interest
(Section~\ref{softwarequalities}), along with information on how we assessed
these qualities. The following section (Section~\ref{Sec_OverallProcess})
provides an overview of the steps we used to select, measure and compare LBM
software.  The remaining sections provide details on each step of the steps in
the methodology. 
% of the next sections provide an overview of the specifics of identifying
% (Section~\ref{identifysoftware}) and filtering (Section~\ref{filtersoftware})
% the candidate software, our quantitative measurements
% (Section~\ref{empiricalmeasures}), the AHP ranking process
% (Section~\ref{AHP}), developer interviews (Section~\ref{SecSurvey}), and
% interactions with one of the key members of the assessment team -- the Domain
% Expert (Section~\ref{sec_vet_software_list}).

@@

We adopt software quality definitions from various researchers and subject
matter experts. The following are the software quality definitions used in this
exercise, along with comments regarding their measurement.

@@

	\item[Installability] Installability is measured by the effort required for
	the installation, uninstallation or reinstallation of a software product in
	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. A good
	measure of installability correlates with scenarios when low or moderate
	effort is required to gather and prepare software for its general use on a
	system for which it was designed. In our case effort includes the time spent
	finding and understanding the installation instructions, the person-time and
	resources spent performing the installation procedure, and the absence or
	ease of overcoming system compatibility issues. An installability score
	improves if the software has a means to validate the installation and
	instructions for uninstallation.

	\item[Correctness] A software program is correct if it behaves according to
	its stated specifications \citep{GhezziEtAl2003}. This requires that the
	specification is available. Scientific software is unlikely to have an
	explicit specification, since this is not common practice in the field. As a
	consequence, the correctness of software cannot often be measured directly.
	We assess correctness indirectly by looking for the following: availability
	of a requirements specification, reference to domain theory, and explicit
	use of tools and/or techniques for building confidence, such as
	documentation generators and software analysis tools.

@@

	\item[Reliability] Reliability is measured by the probability of
	failure-free operation of a computer program in a specified environment for
	a specified time \citep{GhezziEtAl2003, musa1987software}. Reliability is
	thus positively correlated with the absence of errors during installation
	and use. Recoverability from errors also improves reliability.

	\item[Robustness] Software possesses the characteristic of robustness if it
	behaves ``reasonably'' in two situations: i) when it encounters
	circumstances not anticipated in the requirements specification; and ii)
	when the assumptions in its requirements specification are violated
	\citep{boehm2007software, ghezzi1991fundamentals}. A good measure of
	robustness correlates with a reasonable reaction to unexpected input,
	including data of the wrong type, empty input, or missing files or links.
	Examples of reasonable reaction include an appropriate error message and the
	ability to recover the system.

@@

	\item[Maintainability] A measure of maintainability is the effort with which
	a software system or component can be modified to correct faults, improve
	performance or other attributes, and satisfy new requirements
	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work
	maintainability is measured by the quality of documentation artifacts, and
	the presence of version control and issue tracking. These artifacts can
	greatly decrease the effort needed to modify software. There are many
	documentation artifacts that can improve maintainability, including user and
	developer manuals, specifications, README files, change logs, release notes,
	publications, forums, and instructional websites.

	\item[Modifiability] Modifiability refers to the ease with which stable
	changes can be made to a system and the flexibility of the system to adopt
	such changes \citep{8016712}. We did not directly measure modifiability.
	Instead, developers were asked in interviews if they considered the ease of
	future changes when developing the software packages, specifically changes
	to the structure of the system, modules and code blocks. A follow up
	question is asked if any measures had been taken.

	\item[Reusability] Reusability refers to the extent to which components of a
	software package can be used with or without adaptation in other software
	packages \citep{kalagiakos2003non}. A good measure of reusability results
	from a large number of easily reusable components. Increased software
	modularization, defined as the presence of smaller components with well
	defined interfaces, is important. For this state of the practice assessment,
	we assumed that a good measure of reusability correlates with an increased
	number of code files, and the availability of API documentation.

@@

	\item[Traceability] Traceability refers to the ability to link the software
	implementation and the software artifacts, especially the requirement
	specification \citep{McCallEtAl1977}. Similar to the quality of correctness,
	this requires the availability of some form of specification. This quality
	refers to keeping track of information as it changes forms or relates
	between artifacts. We did not quantitatively measure traceability. Instead,
	developers were asked in interviews how documentation fits into their
	development process.

	\item[Visibility and Transparency] Visibility and transparency refer to the
	extent to which all of the steps of a software development process, and the
	current status of it, are conveyed clearly \citep{ghezzi1991fundamentals}.
	In this state of the practice assessment a good measure of visibility and
	transparency correlates with a well defined development process,
	documentation of the development process and environment and software
	package version release notes.

@@

	\item[Unambiguity] Unambiguity refers to the extent to which two readers
	have similar interpretations when reading software artifacts. In other
	words, artifacts are unambiguous if, and only if, they only have one
	interpretation \citep{IEEE1998}. This state of the practice assessment did
	not quantitatively measure unambiguity, but we did ask developers if they
	think that the current documentation can clearly convey all necessary
	knowledge to the users, and how they achieved this or what improvements are
	needed to achieve it.

@@

\begin{enumerate}
	\item List candidate software packages for the domain.
	(Section~\ref{identifysoftware})
	\item Filter the software package list. (Section~\ref{filtersoftware})
	\item Gather the source code and documentation for each software package.
	\item Collect quantitative measures from the project repositories.
	(Section~\ref{empiricalmeasures})
	\item Measure using the measurement template.  The full measurement template
	can be found in \citet{SmithEtAl2021}. (Section~\ref{empiricalmeasures})
	\item Survey the developers. (Section~\ref{SecSurvey})
	\item Use AHP to rank the software packages. (Section~\ref{AHP})
	\item Analyze the results and answer the research questions.
\end{enumerate}

The above steps depend on interaction with a Domain Expert partner, as discussed
in Section~\ref{sec_vet_software_list}.

@@

\begin{enumerate}
	\item The software functionality must fall within the identified domain.
	\item The source code must be viewable.
	\item The empirical measures should be available, which implies a preference
	for GitHub-style repositories.
	\item The software cannot be marked as incomplete or in an initial state of
	development.
\end{enumerate}

@@

For the third item in the above filter, software packages were characterized as
`alive' if their related documentation had been updated within the last 18
month. Packages were categorized as `dead' if the last update of this
information was more than 18 month ago.

@@

Filtering by scope, usage, and age decreased the size of the list to 23
packages. Many of the 22 packages that were removed could not be tested as there
was no installation guide, they were incomplete, source code was not publicly
available, a license was needed, or the project was out of scope or not up to a
standard that would support incorporating them into this study. These eliminated
software packages are listed in the Appendix of \citet{Michalski2021}. Of the
remaining 23 packages that were studied, some were kept on the list despite
being marked as dead due to their prevalence on authoritative lists on LBM
software and due to their surface excellence.  After the initial measurement
exercise, which took place in mid 2020, we added one more package in January
2022: Musubi.  Musubi did not come up in our initial search, but once we learned
of it, we felt that this report is incomplete without including it.  With the
addition of Musubi we did not update the manual collection of data for the
previous packages, due to time constraints, but we did redo all data collection
that was automated by tools.

@@

Tools were use to find some of the measurements, such as the number of files,
number of lines of code (LOC), percentage of issues that are closed, etc. The
tool \href{https://github.com/tomgi/git_stats}{GitStats} was used to measure
each software package's GitHub repository for the number of binary files, the
number of added and deleted lines, and the number of commits over varying time
intervals. The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code
(scc)} was used to measure the number of text based files as well as the number
of total, code, comment, and blank lines in each GitHub repository.

As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
optimal testing environments for each package. VMs were used because it is
easier to start with a fresh environment without having to worry about existing
libraries and conflicts. Moreover, when the tests are complete the VM can be
deleted, without any impact on the host operating system. The most significant
advantage of using VMs is to level the playing field. Every software install
starts from a clean slate, which removes ``works-on-my-computer'' errors.

@@

This project used a tool for conducting this process. The tool includes a
sensitivity analysis that was used to ensure that the software package rankings
are appropriate with respect to the uncertainty of the quality scores. For the
sensitivity analysis we modified the score by 10\% for each package and verified
that the overall ranking was stable.  The
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
file of the tool includes requirements and usage information.

@@

Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints}, \rqref{RQ_Concerns} and
\rqref{RQ_Recommend}) require going beyond the quantitative data from the
measurement template. To gain the required insight, we interviewed developers
using a list of 20 questions from \citet{SmithEtAl2021}. The questions cover the
background of the development teams, the interviewees, and the software itself.
We ask the developers how they organize their projects. We also ask them about
their understanding of the users. Some questions focus on the current and past
difficulties, and the solutions the team has found, or will try. We also discuss
the importance of, and the current situation for, documentation. A few questions
are about specific software qualities, such as maintainability,
understandability, usability, and reproducibility. The interviews are
semi-structured based on the question list; we ask follow-up questions when
necessary. Based on our experience, the interviewees often bring up exciting and
unexpected ideas.

We requested interviews with developers for all 24 packages, except Munubi,
since Munubi was added after our ethics approved had expired.  We were able to
recruit 4 developers to participate in our study.  Our potential interview
subjects were found from project websites, code repositories, publications, and
biographic information on institution web-pages. Meeting were held using on-line
meeting software (either Zoom or Teams), which facilitated recording and
automatic transcription of the meetings.  Each meeting lasted about an hour. The
interview process was approved by the McMaster University Research Ethics Board
under the application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

@@

In this section we answer \rqref{RQ_HighestQuality} by using AHP to rank the
identified LBM software against best practices.  Before discussing quality
differences between the packages and the results of the AHP ranking, we will
contrast the different packages based on their features, as shown in
Table~\ref{tbl_features}. The set of LBM programs form a program family, or a
software product line.  The packages all have the commonality of using LBM
techniques to simulate the motion of fluids, but there are also variabilities
between them.  LBM packages can be distinguished by the dimension of problems
they simulate, whether or not (and how) they support parallel computation,
whether or not compressibility can be considered, whether reflexive boundaries
can be employed, whether the simulation handles multi-fluids, whether the
simulations handles turbulence, and whether complex geometries can be simulated.
All of these variabilities are listed as columns in Table~\ref{tbl_features}
with the software listed in alphabetical order. A more complete analysis of the
commonalities and variabilities between LBM packages is provided in
\citet{Michalski2021}.

The sections below summarize how the software packages compare based on the
qualities discussed in Section~\ref{softwarequalities}, as measured via the
measurement template (Section~\ref{empiricalmeasures}).  For the purpose of
discussion, the last subsection combines all the qualities into one ``global''
ranking.

@@

Tutorials are available for 19 of the software packages. Generally they are
linearly written and easy to follow. However, only eight tutorials provide an
expected output. It is not possible to verify the correctness of the output of
the software packages that are missing this key information. In these cases the
user may need to assume correctness if there are no visible errors.  A
particularly  detailed tutorial that include expected output is provided by
\href{https://www.walberla.net/doxygen/index.html}{waLBerla}.

Unit tests are only explicitly available for two of the software packages,
Ludwig and Musubi. Code modularization of most packages allow for users to
create tests with varying degrees of effort. These tests allow developers and
users to verify the correctness of fragments of the source code, and in doing so
better assess the correctness of the entire package.

@@

The analysis of surface reliability focused on package installation and
tutorials. Errors occurred when installing 16 of the software packages. Every
instance prompted an error message. These messages indicated unrecognized
commands (even when following the installation guide), missing links, missing
dependencies, and syntax errors in code files. In some instances the error
messages were vague. Several automatic installation processes could not find and
load dependencies. In these instances the installation tried to access outdated
external repositories. Seven of the installations were recovered and verified,
and one of the installations (LB3D-Prime) was assumed to be recovered due to the
absence of any way to verify it. The installation of eight of the software
packages could not be recovered. Most of these broken installations could not
find external dependencies, encountered system incompatibilities, or displayed
vague error messages.

Of the 14 software packages that installed correctly and also have tutorials,
four (pyLBM, ESPResSo++, LIMBES, Ludwig) broke during tutorial testing. All of
these instances resulted in an error message being displayed. One error (pyLBM)
was due to a missing tutorial dependency, another (Ludwig) was due to an invalid
command despite following the tutorial, and the final two errors were vague
execution errors. Of the four broken tutorial instances, only the one that was
missing a dependency was recoverable.

@@

Figure~\ref{Fig_Robustness} shows the surface robustness ranking of the software
packages. Software packages with a high score behaved reasonably in response to
unexpected input as described above. All of the software packages that installed
correctly passed this test. They output descriptive error messages or at least
do not crash. Software packages with a lower surface robustness score did not
install correctly, so their robustness score may not be a true reflection of
runtime robustness. All successfully installed software packages that require a
plain text input file correctly handled an unexpected change to these input
files, including a replacement of new lines with carriage returns. Furthermore,
nine of the top 10 ranked packages are noted as being alive.

@@

Software packages were reviewed for the presence of artifacts. Every type of
artifact or file that is not a code file was recorded. The software packages
were also reviewed for software release and documentation version numbers. This
information can be used to troubleshoot issues and organize documentation. All
but three software packages (LatBo.jl, LB3D-Prime, MechSys) have source code
release and documentation version numbers.

@@

Issue tracking is used in 23 software packages, 15 of which use GitHub or GitLab
to host their git repository, seven use email, and one (SunlightLB) uses
SourceForge. Most software packages that use GitHub/GitLab have most of their
issues closed, and only three (laboetie, lettuce, Sailfish) have less than 50
percent of their issues closed. All of the top five overall ranked packages have
most of their issues closed. Alive packages (11 use issue tracking) have 64\% of
their issues closed, while dead packages (3 use issue tracking) have 71\% of
their issues closed. This information is presented in Table~\ref{gitrepodata}.
With respect to the host for the git repository, 13 packages use GitHub and two
(Palabos, waLBerla) use GitLab. Of the other packages, one package (SunlightLB)
uses CVS for issue tracking and version control, and seven packages do not
appear to use any issue tracking system.

@@

Figure~\ref{Fig_Understandability} shows the surface understandability ranking
of the software packages. Software packages with a high score have a consistent
indentation and formatting style, and consistent, distinctive, and meaningful
code identifiers. They also have symbolic constants, and explicitly identify
mathematical and LBM algorithms. Their comments are clear and indicate what is
being done in the source code. The source code is well modularized and
structured. All of the top five ranked packages are noted as being alive.

@@

Software package artifacts were reviewed for the identification of a specific
development model, like a waterfall of agile development model, and the presence
of documentation recording the development process and standard. They were also
reviewed for the identification of the development environment, and the presence
of release notes. The packages tended to not explicitly use well-known
development models. This was also noted in the interviews with developers
(Section~\ref{Sec_CompareMethodologies}). The development teams of these
packages are fairly small and easily organized without the need for such
processes. Seven of the software packages did have some artifacts outlining the
general development process, how to contribute, and the status of the package or
its components. Eight of the packages have artifacts that note the development
environment. While this information could help developers, and would improve
transparency, the small close-knit nature of the development teams make
explicitly publicly specifying this information practically unnecessary. Version
release notes were found in nine of the software packages.

@@

\subsection{Reproducibility} \label{SecReproducibility}

>>>>

\begin{keyword}
	Lattice Boltzmann Method (LBM), research software, software engineering,
	software quality, Analytic Hierarchy Process
\end{keyword}

@@

A small sample of important applications of LBM include the following: designing
fuel cells \citep{ZhangEtAl2018}, modelling groundwater flow
\citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
system \citep{SadeghiEtAl2020}.  As these examples illustrate, LBM can be used
for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety.  Given the important applications of LBM,
users of the libraries will be concerned with software qualities like
reliability, robustness, reproducibility, performance, correctness and
verifiability.  Since their time is valuable, developers that create or modify
LBM libraries will have additional quality concerns, including maintainability,
reusability, and understandability.  With the quality considerations in mind,
and the potentially overwhelming number of choices for existing LBM libraries,
the time is right to assess the state of the practice.  The goal of this report
is to analyze the current state of LBM software development and compare it to
the development practices employed for research software in general.  We want to
highlight success stories that LBM developers can share amongst their community,
and with the broader research software community, while at the same time
looking for areas for potential future improvement.

@@

Only so much information can be gleaned by looking at software repositories.  To
gain additional insight, we need to interview developers.  We need to learn
their concerns, how they deal with these concerns; we need to learn what pain
points exist for them.  We wish to know what practices are used by the top LBM
projects, so that others can potentially emulate these practices. We also wish
to identify new practices that LBM developers can adopt to improve their
software in the future. The above points are covered by the questions outlined
below:

@@

We developed a methodology for evaluating the state of the practice of
research software \citep{SmithEtAl2021}. This methodology can be instantiated
for a specific domain of research software, which in the current case is LBM
software.  Since we group best practices around conventional software qualities,
the first section below provides definitions for the qualities of interest
(Section~\ref{softwarequalities}), along with information on how we assessed
these qualities. The following section (Section~\ref{Sec_OverallProcess})
provides an overview of the steps we used to select, measure and compare LBM
software.  The remaining sections provide details on each step of the steps in
the methodology. 
% of the next sections provide an overview of the specifics of identifying
% (Section~\ref{identifysoftware}) and filtering (Section~\ref{filtersoftware})
% the candidate software, our quantitative measurements
% (Section~\ref{empiricalmeasures}), the AHP ranking process
% (Section~\ref{AHP}), developer interviews (Section~\ref{SecSurvey}), and
% interactions with one of the key members of the assessment team -- the Domain
% Expert (Section~\ref{sec_vet_software_list}).

@@

The following are the software quality definitions used in this exercise, along
with comments regarding their measurement.

@@

	\item[Installability] Installability is measured by the effort required for
	the installation, uninstallation or reinstallation of a software product in
	a specified environment \citep{ISO/IEC25010, lenhard2013measuring}. In our
	case effort includes the time spent finding and understanding the
	installation instructions, the person-time and resources spent performing
	the installation procedure, and the absence or ease of overcoming system
	compatibility issues. An installability score improves if the software has a
	means to validate the installation and instructions for uninstallation.

	\item[Correctness] A software program is correct if it behaves according to
	its stated specifications \citep[p.\ 17]{GhezziEtAl2003}. This requires that
	the specification is available. Research software is unlikely to have an
	explicit specification, since this is not common practice in the field. As a
	consequence, the correctness of software cannot often be measured directly.
	We assess correctness indirectly by looking for the following: availability
	of a requirements specification, reference to domain theory, and explicit
	use of tools and/or techniques for building confidence, such as
	documentation generators and software analysis tools.

@@

	\item[Reliability] Reliability is measured by the probability of
	failure-free operation of a computer program in a specified environment for
	a specified time \citep[p.\ 357]{GhezziEtAl2003}, \citep{musa1987software}.
	Reliability is thus positively correlated with the absence of errors during
	installation and use. Recoverability from errors also improves reliability.

	\item[Robustness] Software possesses the characteristic of robustness if it
	behaves ``reasonably'' in two situations: i) when it encounters
	circumstances not anticipated in the requirements specification; and ii)
	when the assumptions in its requirements specification are violated
	\citep{boehm2007software}, \citep[p.\ 19]{GhezziEtAl2003}. A good measure of
	robustness correlates with a reasonable reaction to unexpected input,
	including data of the wrong type, empty input, or missing files or links.
	Examples of reasonable reactions include an appropriate error message and the
	ability to recover the system.

@@

	\item[Maintainability] A measure of maintainability is the effort with which
	a software system or component can be modified to correct faults, improve
	performance or other attributes, and satisfy new requirements
	\citep{IEEEStdGlossarySET1990, boehm2007software}. In the current work
	maintainability is measured by the quality of the documentation artifacts,
	and the presence of version control and issue tracking. These artifacts can
	greatly decrease the effort needed to modify software. There are many
	documentation artifacts that can improve maintainability, including user and
	developer manuals, specifications, README files, change logs, release notes,
	publications, forums, and instructional websites.

	\item[Modifiability] Modifiability refers to the ease with which stable
	changes can be made to a system and the flexibility of the system to adopt
	such changes \citep{8016712}. We did not directly measure modifiability.
	Instead, developers were asked in interviews if they considered the ease of
	future changes when developing the software packages, specifically changes
	to the structure of the system, modules and code blocks. A follow up
	question is asked if any measures have been taken.

	\item[Reusability] Reusability refers to the extent to which components of a
	software package can be used with or without adaptation in other software
	packages \citep{kalagiakos2003non}. A good indicator of reusability is a
	large number of easily reusable components. Therefore, we look for high
	modularization, which is defined as the presence of smaller components with
	well defined interfaces. For this state of the practice assessment, we
	assume that a good measure of reusability correlates with a high
	number of code files, and the availability of API documentation.

@@

	\item[Traceability] Traceability refers to the ability to link the software
	implementation and the software artifacts, especially the requirement
	specification \citep{McCallEtAl1977}. Similar to the quality of correctness,
	this requires the availability of some form of requirements specification.
	This quality refers to keeping track of information as it changes forms or
	relates between artifacts. We did not quantitatively measure traceability.
	Instead, developers were asked in interviews how documentation fits into
	their development process.

	\item[Visibility and Transparency] Visibility and transparency refer to the
	extent to which all of the steps of a software development process, and the
	current status of it, are conveyed clearly \citep[p.\ 32]{GhezziEtAl2003}.
	In this state of the practice assessment a good measure of visibility and
	transparency correlates with a well defined development process,
	documentation of the development process and environment, and software
	version release notes.

@@

	\item[Unambiguity] Unambiguity refers to the extent to which two readers
	have similar interpretations when reading software artifacts. In other
	words, artifacts are unambiguous if, and only if, they only have one
	interpretation \citep{IEEE1998}. This state of the practice assessment did
	not quantitatively measure unambiguity, but we did ask developers if they
	believe that their current documentation clearly conveys all necessary
	knowledge to the users.  Follow-up questions ask how they achieved
	unambiguity, or what improvements are needed to achieve it.

@@

\begin{enumerate}
	\item List candidate software packages for the domain.
	(Section~\ref{identifysoftware})
	\item Filter the software package list. (Section~\ref{filtersoftware})
	\item Gather the source code and documentation for each software package.
	\item Collect quantitative measures from the project repositories.
	(Section~\ref{empiricalmeasures})
	\item Measure using the measurement template.  The full measurement template
	can be found in \citet{SmithEtAl2021}. (Section~\ref{empiricalmeasures})
	\item Use AHP to rank the software packages. (Section~\ref{AHP})
	\item Interview the developers. (Section~\ref{SecSurvey})
	\item Domain analysis. (Section~\ref{SecDomainAnalysis})
	\item Analyze the results and answer the research questions.
\end{enumerate}

\noindent The above steps depend on interaction with a Domain Expert partner, as
discussed in Section~\ref{sec_vet_software_list}.

@@

\begin{enumerate}
	\item The software functionality must fall within the identified domain.
	\item The source code must be viewable.
	\item The empirical measures should be available, which implies a preference
	for GitHub-style repositories.
	\item The software cannot be marked as incomplete, or in an initial state of
	development.
\end{enumerate}

@@

For the third item in the above filter, software packages were characterized as
`alive' if their related documentation had been updated within the last 18
months. Packages were categorized as `dead' if the last update was more than 18
month ago.

@@

Filtering by scope, usage, and age decreased the size of the list to 23
packages. Many of the 22 packages that were removed could not be tested as there
was no installation guide, they were incomplete, source code was not publicly
available, a license was needed, or the project was out of scope or not up to a
standard that would support incorporating them into this study. These eliminated
software packages are listed in the Appendix of \citet{Michalski2021}. Of the
remaining 23 packages that were studied, some were kept on the list despite
being marked as dead due to their prevalence on authoritative lists on LBM
software and due to their surface excellence.  After the initial measurement
exercise, which took place in mid 2020, we added one more package in January
2022: Musubi.  Musubi did not come up in our initial search, but once we learned
of it, we felt that this report was incomplete without including it.  With the
addition of Musubi we did not update the manual collection of data for the
previous packages, due to time constraints, but we did redo all data collection
that was automated by tools.

@@

Tools were used to find some of the measurements, such as the number of files,
number of lines of code (LOC), percentage of issues that are closed, etc. The
tool \href{https://github.com/tomgi/git_stats}{GitStats} was used to measure
each software package's GitHub repository for the number of binary files, the
number of added and deleted lines, and the number of commits over varying time
intervals. The tool \href{https://github.com/boyter/scc}{Sloc Cloc and Code
(scc)} was used to measure the number of text based files as well as the number
of total, code, comment, and blank lines in each GitHub repository.

As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
optimal testing environment for each package. VMs were used because it is
easier to start with a fresh environment without having to worry about existing
libraries and conflicts. Moreover, when the tests are complete the VM can be
deleted, without any impact on the host operating system. The most significant
advantage of using VMs is to level the playing field. Every software install
starts from a clean slate, which removes ``works-on-my-computer'' errors.

@@

This project used a tool we developed for conducting the AHP analysis. The tool
includes a sensitivity analysis to ensure that the software package rankings are
appropriate with respect to the uncertainty of the quality scores. For the
sensitivity analysis we modified the score by 10\% for each package and verified
that the overall ranking was stable.  The
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/AHP2020/LBM/README.txt}{README}
file of the tool includes requirements and usage information.

@@

Several of the research question (\rqref{RQ_CompareToolsProjMngmnt},
\rqref{RQ_CompareMethodologies}, \rqref{RQ_PainPoints}, \rqref{RQ_Concerns} and
\rqref{RQ_Recommend}) require going beyond the quantitative data from the
measurement template. To gain the required insight, we interviewed developers
using a list of 20 questions from \citet{SmithEtAl2021}. The questions cover the
background of the development teams, the interviewees, and the software itself.
We ask the developers how they organize their projects. We also ask about their
understanding of the users. Some questions focus on the current and past
difficulties, and the solutions the team has found, or will try. We also discuss
the importance of, and the current situation for, documentation. A few questions
are about specific software qualities, such as maintainability,
understandability, usability, and reproducibility. The interviews are
semi-structured based on the question list; we ask follow-up questions when
necessary. Based on our experience, the interviewees often bring up exciting and
unexpected ideas.

We requested interviews with developers for all 24 packages, except Munubi,
since Munubi was added after our ethics approved had expired.  We were able to
recruit 4 developers to participate in our study.  Our potential interview
subjects were found from project websites, code repositories, publications, and
biographic information on institution web-pages. Meeting were held using on-line
meeting software (either Zoom or Teams), which facilitated recording and
automatic transcription.  Each meeting lasted about an hour. The interview
process was approved by the McMaster University Research Ethics Board under the
application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

\subsection{Domain Analysis} \label{SecDomainAnalysis}

Since the LBM domain has a reasonably small scope, we are able to view the
software as constituting a program family.  The concept of a program family is
defined by \citet{parnas1976design} as ``a set of programs whose common
properties are so extensive that it is advantageous to study the common
properties of the programs before analyzing individual members''. Studying the
common properties within a family of related programs is termed a domain
analysis.

The domain analysis (also called a commonality analysis) consists of
understanding the commonalities, variabilities and common terminology for a
program family \citep{weiss1997defining}. Commonalities are goals, theories,
models, definitions and assumptions that are common between family members.
Variabilities are goals, theories, models, definitions and assumptions that
differ between family members. Associated with each variability are its
parameters of variation, which summarize the possible values for that
variability, along with their potential binding time.  The binding time is when
the value of the variability is set.  It could be set as specification time,
build time (when the program is being compiled) or run time (when the code is
executing). As an example, for the current assessment, the packages all have the
commonality of using LBM techniques to simulate the motion of fluids.  For
variabilities, the LBM packages can be distinguished by the dimension of
problems they simulate, whether or not (and how) they support parallel
computation, whether or not compressibility can be considered, whether reflexive
boundaries can be employed, whether the simulation handles multi-fluids, whether
the simulations handles turbulence, and whether complex geometries can be
simulated.

To keep the current presentation focused, we only present the results of our
shallow domain analysis. For the LBM domain a table was constructed that
distinguishes the programs by their variabilities.  In research software the
variabilities are often assumptions, as shown in Table~\ref{tbl_features}. The
variabilities are listed as columns in Table~\ref{tbl_features} with the
software listed in alphabetical order. A more complete analysis of the
commonalities and variabilities between LBM packages is provided in
\citet{Michalski2021}.

@@

In this section we answer \rqref{RQ_HighestQuality} by using AHP to rank the
identified LBM software against best practices.  The sections below summarize
how the software packages compare based on the qualities discussed in
Section~\ref{softwarequalities}, as measured via the measurement template
(Section~\ref{empiricalmeasures}).  For the purpose of discussion, the last
subsection combines all the qualities into one ``global'' ranking.  The full LBM
data is available publicly on Mendeley Data \citep{Smith2022}.

@@

Tutorials are available for 19 of the software packages. Generally they are
linearly written and easy to follow. However, only eight tutorials provide an
expected output. It is not possible to verify the correctness of the output of
the software packages that are missing this key information. In these cases the
user may need to assume correctness if there are no visible errors.  A
particularly  detailed tutorial that include expected output is provided by
\href{https://www.walberla.net/doxygen/index.html} {waLBerla}.

Unit tests are only explicitly available for two of the software packages,
Ludwig and Musubi. Code modularization of most packages allows for users to
create tests with varying degrees of effort. These tests allow developers and
users to verify the correctness of fragments of the source code, and in doing so
better assess the correctness of the entire package.

@@

The analysis of surface reliability focused on package installation and
tutorials. Errors occurred when installing 16 of the software packages. Every
instance prompted an error message. These messages indicated unrecognized
commands (even when following the installation guide), missing links, missing
dependencies, and syntax errors in code files. In some instances the error
messages were vague. Several automatic installation processes could not find or
load dependencies. In these instances the installation tried to access outdated
external repositories. Seven of the installations were recovered and verified,
and one of the installations (LB3D-Prime) was assumed to be recovered due to the
absence of any way to verify otherwise. The installation of eight of the software
packages could not be recovered. Most of these broken installations could not
find external dependencies, encountered system incompatibilities, or displayed
vague error messages.

Of the 14 software packages that installed correctly and also have tutorials,
four (pyLBM, ESPResSo++, LIMBES, Ludwig) broke during tutorial testing. All of
these instances resulted in an error message being displayed. One error (pyLBM)
was due to a missing tutorial dependency, another (Ludwig) was due to an invalid
command despite following the tutorial, and the final two errors were vague
execution errors. Of the four broken tutorial instances, only the missing
dependency one was recoverable.

@@

Figure~\ref{Fig_Robustness} shows the surface robustness ranking of the software
packages. Software packages with a high score behaved reasonably in response to
unexpected input as described above. All of the software packages that installed
correctly passed this test. They output descriptive error messages or at least
did not crash. Software packages with a lower surface robustness score did not
install correctly, so their robustness score may not be a true reflection of
runtime robustness. All successfully installed software packages that require a
plain text input file correctly handled an unexpected change to these input
files, including a replacement of new lines with carriage returns. Furthermore,
nine of the top 10 ranked packages are noted as being alive.

@@

Software packages were reviewed for the presence of artifacts. Every type of
artifact or file that is not a code file was recorded. (The list of artifacts
will be discussed further in Section~\ref{Sec_CompareArtifacts}.)  The software
packages were also reviewed for software release and documentation version
numbers. This information can be used to troubleshoot issues and organize
documentation. All but three software packages (LatBo.jl, LB3D-Prime, MechSys)
have source code release and documentation version numbers.

@@

Issue tracking is used in 23 software packages, 15 of which use GitHub or GitLab
to host their git repository, seven use email, and one (SunlightLB) uses
SourceForge. Most software packages that use GitHub/GitLab have the majority of
their issues closed, and only three (laboetie, lettuce, Sailfish) have less than
50 percent of their issues closed. All of the top five overall ranked packages
have most of their issues closed. Alive packages (11 use issue tracking) have
64\% of their issues closed, while dead packages (3 use issue tracking) have
71\% of their issues closed. This information is presented in
Table~\ref{gitrepodata}. With respect to the host for the git repository, 13
packages use GitHub and two (Palabos, waLBerla) use GitLab. Of the other
packages, one package (SunlightLB) uses CVS for issue tracking and version
control, and seven packages do not appear to use any issue tracking system.

@@

Figure~\ref{Fig_Understandability} shows the surface understandability ranking.
Software packages with a high score have a consistent indentation and formatting
style, and consistent, distinctive, and meaningful code identifiers. They also
have symbolic constants, and explicitly identify mathematical and LBM
algorithms. Their comments are clear and indicate what is being done in the
source code. The source code is well modularized and structured. All of the top
five ranked packages are noted as being alive.

@@

Software package artifacts were reviewed for the identification of a specific
development model, like a waterfall of agile development model, and the presence
of documentation recording the development process and standards. They were also
reviewed for the identification of the development environment, and the presence
of release notes. The packages tended to not explicitly use well-known
development models. This was also noted in the interviews with developers
(Section~\ref{Sec_CompareMethodologies}). The development teams of these
packages are fairly small and easily organized without the need for such
processes. Seven of the software packages did have some artifacts outlining the
general development process, how to contribute, and the status of the package or
its components. Eight of the packages have artifacts that note the development
environment. While this information could help developers, and would improve
transparency, the small close-knit nature of the development teams make
explicitly publicly specifying this information practically unnecessary. Version
release notes were found in nine of the software packages.

@@

\subsection{Prioritize Documentation} \label{Sec_RecommendDoc}
